---
title: "Historical Production Results (Eagle Ford, TX)"
author: "Shu Cong"
date: "March 30, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = './')  # Relative address

knitr::opts_chunk$set(echo = F,
                      warning = F,
                      fig.height = 4,
                      message = F)


options(xtable.comment = FALSE) # Remove comments when generating LATEX code

# Loading packages
pkgs = c('ggplot2', 'reshape2', 'plyr', 'tidyverse', 'lubridate', 'stringr', 'xtable')
inp = pkgs[!pkgs %in% installed.packages()[,1]]
if(length(inp) > 0){
  install.packages(inp)
}
for(p in pkgs){
  require(p, character.only = T)
}
rm(pkgs, inp, p)
```


```{r data_import}
cache = T # Use cache? If FALSE, you need to look at clusters of kmeans and re-redine the centers
read_intv = 10000;

if(cache){
  load('cache.RData')
}else{
  ##### Import data #####
  comp_file = 'EagleFord_Well Meta.txt'
  prod_file = 'EagleFord_Production.txt'; file_encoding = 'UTF-16LE'
  selected_headers = c('api', 'operator_name')
  
  # Load production file
  read.table(prod_file, header = T, sep = '\t', 
             fileEncoding = file_encoding, stringsAsFactors = F,
             colClasses = c(rep('character', 4), rep('numeric', 4))) ->
    dat
  
  api_list = unique(dat$api)
  
  # Load company file
  f = file(comp_file, open = "r", blocking = F, encoding = file_encoding)
  
  f %>% 
    readLines(f, n = 1) %>% 
    str_split('\\t', simplify = T) ->
    header
  
  cols = which(header %in% selected_headers)
  api_col = which(header %in% 'api')
  
  # Read *read_intv* lines per loop and look for completion, if TRUE, break
  comp_api = NULL; count = 0
  while(length(content <- readLines(f, n = read_intv)) > 0){
    content %>% 
      str_split('\\t', simplify = T) %>% 
      subset.matrix(select = cols) %>% 
      data.frame(stringsAsFactors = F) ->
      tmp
    
    names(tmp) = selected_headers
    
    # Filter out api not on the list
    tmp = tmp[tmp$api %in% api_list, ]
    # api_list = api_list[!api_list %in% tmp]
    
    
    comp_api = rbind(comp_api, tmp)
    
    print(str_c(count * read_intv + length(content), ' rows found! ', nrow(tmp), ' match.')); count = count + 1
  }
  
  ### Join two tables
  data = left_join(dat, comp_api, by = 'api')
  
  ##### Clean #####
  data %>% 
    mutate(
      Month = as.Date(date, '%m/%d/%Y'),
      operator_name = str_replace_all(operator_name, '\\"', ''),
      Year = year(Month)
    ) %>% 
    select(
      API = api,
      Index = index,
      Company = operator_name,
      Year = Year,
      Month = Month,
      Oil_Production = volume_oil_formation_bbls
    ) ->
    dat
  
  ##### Clean #####
  dat %>% 
    filter(Year > 2000) ->
    dat
}


```


## Data Clean

Year 1993 has only two rows, which could be a mistake. These two rows are to be removed. Also, there exists some missing values in the column *index* though I don't think this column plays any role in the analysis. Also, the oil production varies from zero to $`r max(dat$Oil_Production)`$, highly skewed. There could be some outliers. However, I'm not going to delete those points due to the lack of oil rig background.


## Problem 1: Determine and summarize historical aggregate oil production by company over time 

See **[Oil Production] Monthly Aggregate by Company.csv** (Monthly total), or **[Oil Production] Annual Aggregate by Company.csv** (Annual summation) for details, here's visualization result.

Data Description:


* Company: Operator
* Month/Year: Timestamp
* Oil_Production_Aggregate: Total oil production during that month/year
* Active_Wells: Number of wells that has **non-zero production**



```{r}
##### Determine and summarize historical aggregate oil production by company over time #####
dat %>% 
  group_by(Company, Month) %>% 
  dplyr::summarise(Oil_Production_Aggregate = sum(Oil_Production),
                   Well_Numbers = length(unique(API)),
                   Active_Wells = length(unique(API[Oil_Production > 0]))) %>% 
  ungroup() %>% 
  arrange(Company, Month) ->
  OPA_m

dat %>% 
  group_by(Company, Year) %>% 
  dplyr::summarise(Oil_Production_Aggregate = sum(Oil_Production),
                   Well_Numbers = length(unique(API)),
                   Active_Wells = length(unique(API[Oil_Production > 0]))) %>% 
  ungroup() %>% 
  arrange(Company, Year) ->
  OPA_y

OPA_m %>% 
  filter(Company %in% head(unique(Company), 5)) %>% 
  mutate(Company = str_c(str_sub(Company, end = 15), '...'),
         log_Aggregate_production = log10(Oil_Production_Aggregate)) %>% 
ggplot() + 
  aes(x = Month, y = log_Aggregate_production, 
      group = Company, color = Company) +
  geom_line() +
  ggtitle('log10 of Monthly Production by Company (Five Samples)') +
  xlim(as.Date('2011-01-01'), max(dat$Month))

write.csv(OPA_m, '[Oil Production] Monthly Aggregate by Company.csv', row.names = F)
write.csv(OPA_y, '[Oil Production] Annual Aggregate by Company.csv', row.names = F)
```

I choose the first 15 companies and plot their log of oil production from 2011 to as-of-date. I use log (base = 10) here to keep them in the same graph.

## Problem 2: Determine and summarize historical average oil production per well by company

See **[Oil Production] Production by Well X Company.csv** for average production per well by company by API.

See **[Oil Production] Production per Well by Company.csv** for monthly weighted average production per well by Company, here's visualization result.

Output Description: 


* Company: Operator
* API: Api number
* sum_per_Well: For this api, what's the total oil production
* Total_Months: How many months in the data for certain api
* P_per_Month: Average (monthly) oil production for this well


```{r}
# ABRAXAS PETROLEUM CORPORATION 42285334510000 85 0
# ACTIVA RESOURCES, LLC 0

dat %>% 
  group_by(Company, API) %>% 
  dplyr::summarise(
    sum_per_Well = sum(Oil_Production),
    Total_Months = n()
    # Producing_Months = sum(Oil_Production > 0)
    ) %>% 
  mutate(
    P_per_Month = sum_per_Well / Total_Months
    # P_per_Production_Month = sum_per_Well / Producing_Months
  ) %>% 
  ungroup() ->
  OPM_m
# OPM_m$P_per_Production_Month[is.nan(OPM_m$P_per_Production_Month)] = 0


OPM_m %>% 
  group_by(Company) %>% 
  dplyr::summarise(
    Total_Production = sum(sum_per_Well),
    Total_Month = sum(Total_Months),
    # Total_Production_Month = sum(Producing_Months),
    P_per_Well_Monthly = mean(P_per_Month)
    # P_per_Active_Well_Monthly = mean(P_per_Production_Month),
    # P_per_Well_Total = sum(sum_per_Well) / n()
  ) ->
  OPM_h


Mean_Production_Monthly = mean(OPM_h$P_per_Well_Monthly)
OPM_h %>% 
  ggplot() +
  aes(x = P_per_Well_Monthly) +
  geom_histogram(bins = 100) + 
  geom_vline(xintercept = Mean_Production_Monthly, color = 'red') +
  ggtitle(str_c('Monthly Production per Well by Company (Histogram, average = ', str_sub(Mean_Production_Monthly, end = 6), ')'))
  
write.csv(OPM_h, '[Oil Production] Production per Well by Company.csv', row.names = F)

write.csv(OPM_m, '[Oil Production] Production by Well X Company.csv', row.names = F)
```

Histogram of monthly production per well (All wells, including zero-production wells) among these companies. The average is $`r Mean_Production_Monthly`$.


## Problem 3: Based on their history, which company or companies are likely to drill the most productive wells going forward?



```{r}
dat %>% 
  group_by(Company, Month) %>% 
  dplyr::summarise(
    Total_Production = sum(Oil_Production),
    Well_Number = length(unique(API)),
    # Active_Well = length(unique(API[Oil_Production > 0])) / Well_Number,
    Production_per_Well = Total_Production / Well_Number
  ) %>% 
  ungroup() ->
  OP
OP$Production_per_Well[is.nan(OP$Production_per_Well)] = 0
write.csv(OP, '[Oil Production] Monthly Production per Well by Company.csv')
```


```{r}
m = 3
tmp = dcast(OP, Month ~ Company, sum, value.var = 'Production_per_Well')
tmp = tmp[tmp$Month > as.Date('2013-01-01'), ]
n = nrow(tmp)
tm = NULL
for(i in 1:(n-m + 1)){
  tm = rbind(tm, 
             c(
               rep(0, (i - 1)),
               rep(1, m),
               rep(0, (n - m - i + 1))
             ))
}
tm  = tm/m
# OP = data.frame(Month = tmp$Month[-(1:(m-1))] ,tm %*% as.matrix(tmp[,-1]))
tmp = tm %*% as.matrix(tmp[,-1])

nclusters = 4
if(!cache){
  re = kmeans(t(tmp[,]), nclusters)
}

ct = c(1,2)
candidate = names(re$cluster)[re$cluster %in% ct];

tmp = dcast(OP, Month ~ Company, sum, value.var = 'Production_per_Well')
tmp = tmp[, candidate]

apply(tmp, 2, function(x){
  x = tail(x, 6)
  c(Q25 = quantile(x, 0.25),Median = median(x),
    Mean = mean(x),
    Q75 = quantile(x, 0.75), Max = max(x))
}) %>% 
  t() %>% 
  as.data.frame() %>% 
  cbind(Company = candidate) %>% 
  arrange(desc(Median)) ->
  tmp

candidate_table = tmp

```


Top 3 Companies:

1. `r candidate_table$Company[1]`
2. `r candidate_table$Company[2]`
3. `r candidate_table$Company[3]`

Reason: They have upward-trend production per well for the last 3 years and good production per well rate for the lastest 6 months.

See **[Oil Production] Monthly Production per Well by Company.csv** for monthly production per well by company

```{r}
col = c('black', 'green', 'blue', 'purple', 'yellow')[1:nclusters]
matplot(t(re$centers), type='l', col = col, lty = 1,
        xlab = 'Months, from 2013-01-01',
        ylab = 'Monthly production per well',
        main = str_c(m, '-Month moving average clusters (Kmeans)')); legend('topleft', str_c('Cluster', 1:nclusters), col = col, lty = 1)
```

The production per well varies too much from month to month. Therefore I calculate the $`r m`$ month moving average and use kmeans learning method to classify these production curve. From the plot above, we might want those in cluster 1, 2

```{r, results='asis'}

xtable(head(candidate_table, 20))


```

There are only `r length(candidate)` companies in these cluster. Therefore I list their last half-year production summaries by calculating 25%, 50%, 75% quantiles as well as average and maximum, sorted by **Median** in decreasing order. (Only select top 20 companies)

```{r}
OP %>% 
  filter(Company %in% candidate_table$Company[1:3],
         Month >= as.Date('2014-08-01')) %>% 
  ggplot()+
  aes(x = Month, y = Production_per_Well, group = Company)+
  geom_smooth(method = 'loess', alpha = 0.3, span = 0.5, size = 1, linetype = "dashed")+
  geom_line(size = 1) +
  facet_wrap(~Company , scale = 'free_x') +
  ggtitle('Production per Well by Month for Top-Three Companies')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

```{r}
if(!cache){
  save(dat, re, file = 'cache.RData')
}

```


